---
title: "Carlos_Parra MovieLens Project"
author: "Carlos Parra"
date: "10/21/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Movie Ratings Prediction
## Introduction
This docouments uses the code provided in the _Capstone section_ to generate the training (edx) and test (validation) datasets obtained from Movielens 10M dataset.
Next libraries will be used to generate the rating
* library(tidyverse)
* library(caret)
* library(data.table)
## Movie Prediction
In order to optimize the movie prediction, I will combine several Machine Learning algorithms  learned during the training  to optimize the result (minimize **root mean squared error- RMSE**)
1. Baseline Preducors
2. Regularization


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
